---
title: "p8105_hw2_nc2710"
author: "Nicole Comfort"
date: "10/2/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```

## Problem 1

This problem focuses on NYC transit data; in particular, information related to each entrance and exit for each subway station in NYC.

*Read and clean the data; retain line, station name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (the ifelse or recode function may be useful).*

*Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?*

```{r import data}
NYC_transit_data = read_csv(file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>% # clean column names
  select(line:entry, vending, ada) %>% # retain only specified information
  mutate(entry_logical = recode(entry,
                       "YES" = TRUE,
                       "NO" = FALSE)) %>% # convert entry from character to logical variable
  select(-entry) # remove old entry character variable
```

The code above imports data regarding NYC Transit information related to each entrance and exit for each subway station in NYC. After importing the file using a relative path, I cleaned the names and selected the following variables to retain in the dataset: line, station name, station's latitude / longitude, routes served, entry, vending (YES/NO for whether it is present), entrance type (Elevator, stair, easement, etc...), and ADA compliance (YES/NO). I then created a new variable using the recode funtion, named "entry_logical". This step converted the "entry" variable from a character (YES/NO) to a logical variable (TRUE/FALSE). I next removed the "entry" character variable from the dataset. This data cleaning process results in a dataframe composed of `r nrow(NYC_transit_data)` rows and `r ncol(NYC_transit_data)` columns. 

However, these data still are not "tidy" because they are in wide format where we have multiple variables for the routes from route1:route11 (and thus a lot of missing data); In reality, "route" itself should be a variable and the route number (1-11) assigned as a value. This would convert our data from wide to long format. 

*Answer the following questions using these data*

* How many distinct stations are there?

* Note that stations are identified both by name and by line (e.g. 125th St A/B/C/D; 125st 1; 125st 4/5); the distinct function may be useful here.

* How many stations are ADA compliant?

* What proportion of station entrances / exits without vending allow entrance?

* Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?



```{r}

```




