---
title: "p8105_hw2_nc2710"
author: "Nicole Comfort"
date: "10/2/2018"
output: github_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

```

## Problem 1

This problem focuses on NYC transit data; in particular, information related to each entrance and exit for each subway station in NYC.

*Read and clean the data; retain line, station name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (the ifelse or recode function may be useful).*

*Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?*

```{r import data}

NYC_transit_data = read_csv(file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") %>%
  janitor::clean_names() %>% # clean column names
  select(line:entry, vending, ada) %>% # retain only specified information
  mutate(entry_logical = recode(entry,
                       "YES" = TRUE,
                       "NO" = FALSE)) %>% # convert entry from character to logical variable
  select(-entry) # remove old entry character variable

```

The code above imports data regarding NYC Transit information related to each entrance and exit for each subway station in NYC. After importing the file using a relative path, I cleaned the names and selected the following variables to retain in the dataset: line, station name, station's latitude / longitude, routes served, entry (whether entry is allowed), vending (YES/NO for whether it is present), entrance type (Elevator, stair, easement, etc...), and ADA compliance (YES/NO). I then created a new variable using the recode funtion, named "entry_logical". This step converted the "entry" variable from a character (YES/NO) to a logical variable (TRUE/FALSE). I next removed the "entry" character variable from the dataset. This data cleaning process results in a dataframe composed of `r nrow(NYC_transit_data)` rows and `r ncol(NYC_transit_data)` columns. 

However, these data still are not "tidy" because they are in wide format where we have multiple variables for the routes from route1:route11 (and thus a lot of missing data); In reality, "route" itself should be a variable and the route number (1-11) assigned as a value. This would convert our data from wide to long format. 

*Answer the following questions using these data*

* How many distinct stations are there?

There are `r nrow(distinct(NYC_transit_data, line, station_name))` distinct stations. 

* How many stations are ADA compliant?

Of those distinct stations, `r mean(NYC_transit_data$ada)` are ADA compliant. 

* What proportion of station entrances / exits without vending allow entrance?

**Entrance/exit the same??**

The proportion of station entrances / exits without vending that allow entrance is 0.3770492.

```{r answer questions}

# how many distinct stations are there: 
nrow(distinct(NYC_transit_data, line, station_name))

# how many of those distinct stations are ADA compliant: 
mean(NYC_transit_data$ada)

# proportion of station entrances / exits without vending that allow entrance:
no_vending = filter(NYC_transit_data, vending == "NO") # filter to examine only entrances/exits without vending
mean(no_vending$entry_logical) # take the proportion of those that do allow entrance (entry_logical = TRUE)

```

*Reformat data so that route number and route name are distinct variables.* 

**Does this look right?**

```{r}

NYC_transit_data_long =
  gather(NYC_transit_data, key = route_number, value = route_name, route1:route11) # reformat using gather function 

```

*How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?*

60 distinct stations serve the A train. **I don't think that's right... it says some like 168 that I know serve more than A. Do I have to do ifelse?** Of the stations that serve the A train, 17 are ADA compliant. 

```{r}

# how many distinct stations serve the A train: 
A_line =
  filter(NYC_transit_data_long, route_name == "A") %>% # filter to examine only stations serving A
  distinct(., line, station_name, .keep_all = TRUE) # how many distinct stations (line and station) serve A

nrow(A_line) # gives the number of distinct stations that serve the A line

# how many of these are ADA compliant:
filter(NYC_transit_data_long, route_name == "A") %>% # filter to examine only stations serving A
  distinct(., line, station_name, .keep_all = TRUE) %>% # how many distinct stations (line and station) serve A
  filter(., ada == TRUE) %>% # now see how many of these are ADA compliant 
  nrow() # number of rows gives the number ADA compliant

```


## Problem 2 

This problem uses the Mr. Trash Wheel dataset. 

Read and clean the Mr. Trash Wheel sheet:

```{r import Mr. Trash Wheel data}

library(readxl)
trash_data = read_excel("./data/HealthyHarborWaterWheelTotals2017-9-26.xlsx") %>% 
  janitor::clean_names()

```


specify the sheet in the Excel file and to omit columns containing notes (using the range argument and cell_cols() function)
use reasonable variable names
omit rows that do not include dumpster-specific data
rounds the number of sports balls to the nearest integer and converts the result to an integer variable (using  as.integer)
Read and clean precipitation data for 2016 and 2017. For each, omit rows without precipitation data and add a variable year. Next, combine datasets and convert month to a character variable (the variable month.name is built into R and should be useful).

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in both resulting datasets, and give examples of key variables. For available data, what was the total precipitation in 2017? What was the median number of sports balls in a dumpster in 2016?



